A comparison on R tools for nonlinear least squares modeling

### 2023-05-03 RefactoringNLS/ComparingNLStools.Rmd
update with some simplifications

In reality, there are many problems were $D$ should not be ignored, but the work
to compute it precisely is considerable. Many workarounds have been 
proposed, of which the Levenberg-Marquardt (@Levenberg1944, @Marquardt1963) is
the most commonly used. Henceforth, we will simply say "Marquardt", as we believe,
was the first to actually incorporate the ideas into a practical computer
program.

The suggestion is that $D$ be replaced by a multiple of the unit matrix or else
a multiple of the diagonal part of $J' J$. For machines with exceptionally low
precision arithmetic found on some minicomputers in the 1970s, one of us suggested
a linear combination of these ideas. In low precision, some elements of 
$J' J$ could underflow to zero (@jn77ima). We note that the various choices for
$D$, especially the scaling multiple, as well as possible line search along the
direction $\delta$ rather than a unit step (@Hartley1961), 
give rise to a panoply of variant
algorithms. "Marquardt's method" is a family of methods. Fortunately, most choices
work well.

@@

Let us specify in R the three model formulae and set some starting
values for parameters. These starting points are NOT equivalent and
are deliberately crude choices. Workers performing many calculations 
of a similar nature should try to provide good starting points to reduce
computation time and the chance a false solution is found.

@@

One of the useful features of `nls()` is the possibility of a `selfStart` model,
where starting parameter values are not required. We discuss aspects of this
feature at several points below. However, if a `selfStart` model 
is not available, `nls()` sets all the starting parameters 
to 1. This is, in our view, tolerable, but could be improved by using a set of values
that are all slightly different, which, in the case of the example 
model $y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)$
would avoid a singular Jacobian because $b$ and $d$ were equal in value. 
Program modifications to give a sequence like  1.0, 1.1, 1.2, 1.3 for the four 
parameters are fairly obvious.

@@

```{r ex02minpack}
library(minpack.lm)
unlm1<-try(nlsLM(formula=frmu, start=stu1, data=weeddf))
faraway::sumary(unlm1) # Use summary() to get display
# unlm1 # Note the difference. Use this form to get sum of squares
# pnlm0(unlm1)  # Short form of output
snlm1<-try(nlsLM(formula=frms, start=sts1, data=weeddf))
pnlm0(snlm1)
# summary(snlm1)
## Following structure needed to deal with M1mac issue
tnlm1<-try(nlsLM(formula=frmt, start=stt1, data=weeddf))
if (inherits(tnlm1, "try-error")) {
   cat("Failure to compute solution -- likely singular Jacobian\n")
} else {  
   pnlm0(tnlm1) # short form to give sum of squares, else use summary(tnlm1)
}   
```

@@

The output object of `nlxb()` is smaller than the class `nls` object returned
by `nls()`, `nlsLM()` and `gsl_nls()`. Package `nlsr` emphasizes the solution 
of the nonlinear least squares problem rather than the estimation of a nonlinear 
model that fits or explains the data. `nls()` and `nlsLM` return an object of 
class `nls`, which allows for a number of specialized modeling and diagnostic extensions. For compatibility, 
the `nlsr` package has function `wrapnlsr()`, for which `nlsr()` is an alias. 
This uses `nlxb()` to find good parameters, then calls `nls()` to return the class 
`nls` object. Unless particular modeling features are needed, the use of 
`wrapnlsr()` is unnecessary and wasteful of resources.

@@

Termination after what may be considered excessive computation is also important.
`nls()` stops after `maxiter` "iterations". The meaning of "iteration" may require
examination of the code. We prefer to use the number of times the residuals or 
the jacobian have been computed and put upper limits on these. Our codes exit (terminate) 
when these limits are reached. Generally we prefer larger limits than the 
default `maxiter = 50` of `nls()`, avoiding the risk of stopping early.

@@

@@

In considering `selfStart` models, we noted that the base-R function `SSlogis`
is intended to solve our example problem **Logistic3T**.  This example is one 
where the `getInitial()` function (used to find
starting values) actually calls `nls()` recursively with a non-default algorithm
to get "starting" parameters. Then it returns to running the default algorithm
unnecessarily. Such circular references are, in our view, prone to creating errors
in code maintenance. 
To provide simpler starting parameters, the function `SSlogisJN`
is now part of package `nlsr`.

@@

### Feature: na.action

@@

?? Arkajyoti: I think this example may get too long or be too unclear.

@@

>>>>

In reality, there are many problems were $D$ should not be ignored, but the work
to compute it precisely is considerable. Many workarounds have been 
proposed, of which the Levenberg-Marquardt stabilization 
(@Levenberg1944, @Marquardt1963) is
the most commonly used. We will use "Marquardt", as we believe he
first incorporated the ideas into a practical computer program.

The suggestion is that $D$ be replaced by a multiple of the unit matrix or else
a multiple of the diagonal part of $J' J$. In low precision, some elements of 
$J' J$ could underflow to zero (@jn77ima), and a linear combination of both
choices was suggested as a compromise. Various choices for
$D$, as well as a possible line search along the
direction $\delta$ rather than a unit step (@Hartley1961), 
give rise to a panoply of variant
algorithms. "Marquardt's method" is a family of methods. Fortunately, most choices
work well.

@@

Let us specify in R the three model formulas and set some starting
values for parameters. These starting points are NOT equivalent and
are deliberately crude choices. Workers performing many calculations 
of a similar nature should try to provide good starting points to reduce
computation time and the chance a false solution is found.

@@

One of the useful features of `nls()` is the possibility of a `selfStart` model,
where starting parameter values are not required. We discuss aspects of this
feature at several points below. However, if a `selfStart` model 
is not available, `nls()` sets all the starting parameters 
to 1. This is tolerable, but could be improved by using a set of values
that are all slightly different, which, in the case of the example 
model $y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)$
would avoid a singular Jacobian because $b$ and $d$ were equal in value. 
Program modifications to give a sequence like  1.0, 1.1, 1.2, 1.3 for the four 
parameters are fairly obvious.

@@

```{r ex02minpack}
library(minpack.lm)
# ??not needed -- library(nlsr) # for pnlm0
unlm1<-try(nlsLM(formula=frmu, start=stu1, data=weeddf))
# summary not exported by faraway??
# faraway::summary(unlm1) # Use summary() to get display
# unlm1 # Note the difference. Use this form to get sum of squares
pnlm0(unlm1)  # Short form of output
snlm1<-try(nlsLM(formula=frms, start=sts1, data=weeddf))
pnlm0(snlm1)
# summary(snlm1)
## Following structure needed to deal with M1mac issue
tnlm1<-try(nlsLM(formula=frmt, start=stt1, data=weeddf))
if (inherits(tnlm1, "try-error")) {
   cat("Failure to compute solution -- likely singular Jacobian\n")
} else {  
   pnlm0(tnlm1) # short form to give sum of squares, else use summary(tnlm1)
}   
```

@@

The output object of `nlxb()` is smaller than the class `nls` object returned
by `nls()`, `nlsLM()` and `gsl_nls()`. Package `nlsr` emphasizes the solution 
of the nonlinear least squares problem rather than the estimation of a nonlinear 
model that fits or explains the data. The object of class `nls` allows for 
a number of specialized modeling and diagnostic extensions. For compatibility, 
the `nlsr` package has function `wrapnlsr()`, for which `nlsr()` is an alias. 
This uses `nlxb()` to find good parameters, then calls `nls()` to return the class 
`nls` object. Unless particular modeling features are needed, the use of 
`wrapnlsr()` is unnecessary and wasteful of resources.

@@

Termination after what may be considered excessive computation is also important.
`nls()` stops after `maxiter` "iterations". The meaning of "iteration" may require
examination of the code. We prefer to use the number of times the residuals or 
the jacobian have been computed and put upper limits on these. Our codes exit (terminate) 
when these limits are reached. Generally we prefer larger limits than the 
default `maxiter = 50` of `nls()` to avoid stopping early.

@@

?? Can we simplify or suggest use of `str()` ??

@@

In considering `selfStart` models, we noted that the base-R function `SSlogis`
is intended to solve our example problem **Logistic3T**.  This example is one 
where the `getInitial()` function (used to find
starting values) actually calls `nls()` recursively with a non-default algorithm
to get "starting" parameters. Then it returns to running the default algorithm
unnecessarily. Such circular references are, in our view, prone to creating errors
in code maintenance. 
To provide simpler starting parameters, the function `SSlogisJN`
is now part of package `nlsr`, but is most useful for `nls()`.

@@

### Feature: na.action (missing value treatment)

@@

<!-- ?? Arkajyoti: I think this example may get too long or be too unclear. -->

@@

<!-- Ben Bolker mentioned a possibility of using a nonlinear model in  -->
<!-- ??I think?? nlme with some sort of linear model form of the parameters. -->
<!-- May need to mention. -->
